---
title: "CSCI E-63C: Week 3 Assignment"
author: "Tony McGovern"
date: "19 September 2017"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
# List of packages required for this analysis
pkg <- c("devtools", "knitr", "tidyr", "readr", "dplyr", "skimr", "ggplot2", "corrplot", "scales", "GGally", "modelr")

# Check if packages are not installed
# Assign not installed packages to new.pkg
new.pkg <- pkg[!(pkg %in% installed.packages())]

# Install any packages not installed
if (length(new.pkg)) {
  install.packages(new.pkg, repos = "http://cran.rstudio.com")
}

# load packages
library(devtools)
library(knitr)
library(tidyr)
library(readr)
library(dplyr)
library(skimr)
library(ggplot2)
library(corrplot)
library(scales)
library(GGally)
library(modelr)

rm(list=ls())

knitr::opts_chunk$set(echo = TRUE)

```

# Problem 1: model of age and shell weight

In this problem, I load, model, visualize, and comment on data that predicts abalone snail age as a function of other physical attributes.

``` {r abalone_load, echo = TRUE }

# load data
url <- "/Users/Luxive/OneDrive/Software_Engineering/Harvard/CSCI E-63c/HW/abalone.data"
abalone.raw <- read_delim(url, ",", col_names = c("sex","length","diameter","height","weight.whole","weight.shucked","weight.viscera","weight.shell","rings"))

```

Describe the data set attributes:
```{r abalone_attributes, echo = FALSE}

# name the data set attributes
attributes <- data.frame(
  "col_name" = c(
    "sex",
    "length",
    "diameter",
    "height",
    "weight.whole",
    "weight.shucked",
    "weight.viscera",
    "weight.shell",
    "rings"), 
  "attribute_name" = c(
    "Sex",
    "Length",
    "Diameter", 
    "Height",
    "Whole Weight",
    "Shucked Weight",
    "Viscera Weight",
    "Shell Weight",
    "Rings"
  ))

kable(attributes)

```


The `abalone` data set has `r dim(abalone.raw)[1] ` rows or observations and `r dim(abalone.raw)[2]` columns or variables. The predictor variables are _sex_, _length_, _diameter_, _height_, _weight.whole_, _weight.shucked_, _weight.viscera_, and _weight.shell_. The response variable is _rings_. The _sex_ variable is a categorical dimension with three distinct values: "F", "I", and "M". All other variables are continuous variables.

## Calculate correlations between all continuous attributes in the data set

### Pearson correlation

```{r correlation_matrix_pearson, echo = TRUE }

# convert 'sex' variable from a character to a factor
abalone.raw$sex <- factor(abalone.raw$sex)

# convert 'rings' variable from an int to a numeric
abalone.raw$rings <- as.numeric(abalone.raw$rings)

# create 'age' variable which represents the snail age equal to rings + 1.5 per data set description
# drop 'rings' variable
abalone.age <- abalone.raw %>% mutate( age = rings + 1.5 ) %>% select( -c(rings) )

# drop categorical 'sex' variable
abalone.cont <- abalone.age %>% select(-c(sex))

# create Pearson correlation matrix
correlations.pearson <- cor(abalone.cont, method = c("pearson"))
kable(correlations.pearson)

```
```{r correlation_plot_pearson, echo = TRUE }

# visualize pearson correlation matrix
corrplot(correlations.pearson, order = "hclust")

```

### Spearman correlation

```{r correlation_matrix_spearman, echo = TRUE }

# create Spearman correlation matrix
correlations.spearman <- cor(abalone.cont, method = c("spearman"))
kable(correlations.spearman)

```

```{r correlation_plot_spearman, echo = TRUE }

# visualize spearman correlation matrix
corrplot(correlations.spearman, order = "hclust")

```

```{r linear_model, echo = TRUE }

#Fit linear model of age as outcome and shell weight as predictor using R function lm, 
model.linear <- lm(age ~ weight.shell, data = abalone.age)

#display the result using summary function, use its output to answer the following questions:
summary(model.linear)

```

The `weight.shell` predictor is significant at both the 5% and 1% levels, which suggests that this variable explains a significant amount of variability in the `age` response variable.

The model $RSE$ is `r summary(model.linear)$sigma`. The model $R^2$ is `r summary(model.linear)$r.squared`.

The y-intercept means that for an average shell weight of zero grams, the snail age would be about `r round(model.linear$coefficients[1],1)` years. This isn’t particularly meaningful as abalone snails should have a non-zero shell weight. The y-intercept is merely an adjustment constant that serves to adjust the line height of the linear model.

The coefficient for the `weight.shell` predictor variable is `r round(model.linear$coefficients[2],4)`, which means that for every one gram increase in shell weight, the `age` variable changes by about `r round(model.linear$coefficients[2],1)` years.

## Create scatterplot of age and shell weight
```{r scatterplot, echo = TRUE }

#Create scatterplot of age and shell weight and add regression line from the model to the plot
ggplot( data = abalone.age, aes( x = weight.shell, y = age )) + 
  geom_point( aes(color = sex )) + 
  geom_smooth( method="lm", formula = y ~ x ) +
  labs(
    title = "Scatter Plot of Abalone Shell Weight by Age, Grouped by Sex",
    subtitle = "Regressing Shell Weight on Age",
    x = "Shell Weight (in grams)",
    y = "Age (in years)",
    color = "Snail Sex"
  ) +
  theme_linedraw()

```


```{r linear_diagnostic_plot, echo = TRUE }

opar <- par(no.readonly=TRUE)
par(mfrow=c(2,2))

#Create diagnostic plots of the model
plot(model.linear)

```

Comment on any irregularities that they present. 
+For instance, does plot of residuals vs. fitted values suggest presence of non-linearity that remained unexplained by the model?
+How does it compare to the plot of the predictor and outcome with regression line added to it that was generated above?

```{r confidence_intervals, echo = TRUE }

#Use function confint to obtain confidence intervals on model parameters
confint(model.linear)

```

```{r linear_prediction, secho = TRUE }

#Use this model and predict function to make predictions for shell weight values of 0.1, 0.2 and 0.3. 
#Use confidence and prediction settings for parameter interval in the call to predict to obtain confidence and prediction intervals on these model predictions. 
predict(model.linear, data.frame(weight.shell = c(0.1, 0.2, 0.3)), interval = 'confidence')
predict(model.linear, data.frame(weight.shell = c(0.1, 0.2, 0.3)), interval = 'prediction')

```

Explain the differences between interpretation of:

+confidence intervals on model parameters and model predictions
+confidence and prediction intervals on model predictions
+Comment on whether confidence or predicion intervals (on predictions) are wider and why


# Problem 2
```{r model_log, echo = TRUE}

#Use lm() to fit a regression model of log-transformed age as linear function of log-transformed shell weight and use summary to evaluate its results. 

# create new tibble with log-transformed age and shell weight
abalone.log.shell <- abalone.age %>%
  mutate(lage = log2(age), lshell = log2(weight.shell))

model.shell.log <- lm(lage ~ lshell, data = abalone.log.shell)

summary(model.shell.log)

```
  
Can we compare fits obtained from using untransformed (above) and log-transformed attributes? 
Can we directly compare $RSE$ from these two models? 
What about comparing $R2R2$? What would we conclude from this? (Please consult ISLR Ch.3.1.3 if unsure) 
What would be the physical meaning of model coefficients this time? 
What does model intercept represent in this case, for example? 
How sensible is this and how does it compare to that from the fit on untransformed data?


```{r scatterplot_log}

#Create a XY-scatterplot of log-transformed predictor and response and add corresponding regression line to it.
ggplot( data = abalone.log.shell, aes( x = lshell, y = lage )) + 
  geom_point( aes(color = sex )) + 
  geom_smooth( method="lm", formula = y ~ x ) +
  labs(
    title = "Scatter Plot of Abalone Shell Weight by Age, Grouped by Sex",
    subtitle = "Regressing Log-Transform of Shell Weight on Age",
    x = "log(Shell Weight)",
    y = "log(Age)",
    color = "Snail Sex"
  ) +
  theme_linedraw()

```

Compare it to the same plot but in untransformed coordinates obtained above.
What would you conclude from such comparison?

```{r log_diagnostic_plot, echo = TRUE}

opar <- par(no.readonly=TRUE)
par(mfrow=c(2,2))

#Make diagnostic plots for model fit on log-transformed age and shell weight.
plot(model.shell.log)

```



Compare their appearance to that for the model using original scale of measurements.
What would you conclude from this comparison about their relative quality?

```{r multiple_model_log, echo = TRUE}

#To explore effects of adding another variable to the model, continue using log-transformed attributes and fit a model of log-transformed age as a function of shell weight and shucked weight (both log-transformed also). 
abalone.log.shucked <- abalone.log.shell %>%
  mutate(lshucked = log2(weight.shucked))

model.shucked.log <- lm(lage ~ lshell + lshucked, data = abalone.log.shucked)

```

```{r multiple_model_summary, echo = TRUE}

#Just an additive model – no interaction term is necessary at this point. 
#Please obtain and evaluate the summary of this model fit, confidence intervals on its parameters and its diagnostic plots. 
summary(model.shucked.log)

```

```{r multiple_model_diagnostics, echo = TRUE}

opar <- par(no.readonly=TRUE)
par(mfrow=c(2,2))

#Where applicable, compare them to the model obtained above and reflect on pros and cons of including shucked weight as another variable into the model.
plot(model.shucked.log)


```