---
title: 'CSCI E-63C: Week 2 Q&A session'
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
library(ggplot2)
library(ISLR)
library(reshape2)
knitr::opts_chunk$set(echo = TRUE)
```

# Questions

* Week 1 homework
* Week 1 lecture
* Upcoming homework
* Upcoming quiz

```{r}
dfTmp <- data.frame(xvals=rnorm(30),yvals=runif(30),nn=rep(c(5,10,20),10))
ggplot(dfTmp,aes(x=xvals,y=yvals,colour=factor(nn)))+geom_point()+facet_wrap(~nn)
ggplot(dfTmp[dfTmp$nn!=5,],aes(x=xvals,y=yvals,colour=factor(nn)))+geom_point()+facet_wrap(~nn)
```


# Factors and type conversion example

```{r typeconversion}
x <- 1:5; someLetters <- letters[seq_along(x)]; someLetters
names(x) <- someLetters; x
names(x) <- sample(names(x)); x
x[someLetters]
fSomeLetters <- factor(someLetters); fSomeLetters
as.character(fSomeLetters); x[as.character(fSomeLetters)]
x[fSomeLetters]
```

# Some of the R conveniences

* `ggplot2` -- popular plotting package providing unified interface to variety of visualizations (e.g. don't have to remember how to combine `boxplot` and `stripchart`)
* tidy data: `reshape2`, `plyr`, ... -- very convenient together with `ggplot`, standardizes many common data transformations
* OOP: class is just an attribute that is easy to change, s3/s4/rc frameworks, dots in names, http://adv-r.had.co.nz/OO-essentials.html
* `shiny` -- enabling interactive R over web

# S3 OOP in R: a word of caution

```{r s3oop}
fSomeLetters
class(fSomeLetters)
is.object(fSomeLetters) & !isS4(fSomeLetters)
class(fSomeLetters) <- "Let's get wild here"; class(fSomeLetters)
fSomeLetters
class(fSomeLetters) <- c("One class","another class"); fSomeLetters
# attribute "class" determines which "print" is called:
class(fSomeLetters) <- "factor"; fSomeLetters
```



# ISLR Ch.2.3.4 (Lab)

```{r ch234}
##?Auto
class(Auto)
dim(Auto)
summary(Auto)
head(Auto)
names(Auto)
colnames(Auto)
# ISLR says there are 397 rows, five of which include ?'s:
for ( tmpCnm in colnames(Auto) ) {
  cat(tmpCnm,class(Auto[,tmpCnm]),sum(Auto[,tmpCnm]=="?"),fill=TRUE)
}
#?write.table
#?read.table
write.table(Auto,"Auto.txt")
autoDat <- read.table("Auto.txt")
head(autoDat)
dim(autoDat)
dim(na.omit(autoDat))
if ( FALSE ) {
  plot(cylinders, mpg)
}
plot(Auto$cylinders, Auto$mpg)
attach (Auto)
plot(cylinders, mpg)
detach (Auto)
plot(Auto$cylinders, Auto$mpg)
ggplot(Auto,aes(x=cylinders,y=mpg))+geom_point()
plot(factor(Auto$cylinders), Auto$mpg, col ="red", varwidth=T, xlab="cylinders", ylab="MPG")
ggplot(Auto,aes(x=factor(cylinders),y=mpg))+geom_point()
ggplot(Auto,aes(x=factor(cylinders),y=mpg))+geom_point()+geom_boxplot()
ggplot(Auto,aes(x=factor(cylinders),y=mpg))+geom_boxplot()+geom_point()
hist(Auto$mpg)
hist(Auto$mpg,col="red")
ggplot(Auto,aes(x=mpg))+geom_histogram()
ggplot(Auto,aes(x=mpg))+geom_histogram(fill="lightgreen",colour="black")
pairs(Auto)
library(GGally)
ggpairs(Auto[,-ncol(Auto)])
summary(Auto)
summary(Auto$mpg)
```

# ISRL Ch.2.4 (Exercises -- applied)

## Problem 8

```{r ch24problem8}
summary(College)
head(College)
pairs(College[,1:10])
ggpairs(College[,1:10])
plot(College$Private,College$Outstate)
ggplot(College,aes(x=Private,y=Outstate)) + geom_boxplot()
ggplot(College,aes(x=Top10perc))+geom_histogram()
College$Elite <- factor(c("No","Yes")[as.numeric(College$Top10perc>50)+1])
summary(College$Elite)
table(College$Elite)
old.par <- par(mfrow=c(1,2),ps=16)
plot(College$Elite,College$Outstate,xlab="Elite")
plot(College$Private,College$Outstate,xlab="Private")
par(old.par)
head(melt(College[,c("Elite","Private","Outstate")]))
head(melt(College[,c("Elite","Private","Outstate")],id="Outstate"))
ggplot(melt(College[,c("Elite","Private","Outstate")],id="Outstate"),aes(x=value,y=Outstate))+geom_boxplot()+facet_wrap(~variable)
ggplot(melt(College[,c("Elite","Private","Outstate")]),aes(x=Private,y=value,fill=Elite,colour=Elite))+geom_boxplot()+facet_wrap(~variable)
ggplot(melt(College[,c("Elite","Private","Outstate")]),aes(x=Private,y=value,fill=Elite,colour=Elite))+geom_boxplot(fill="white",outlier.colour=NA)+geom_point(position=position_jitterdodge())+facet_wrap(~variable)
table(College[,c("Elite","Private")])
fisher.test(table(College[,c("Elite","Private")]))
chisq.test(table(College[,c("Elite","Private")]))
summary(lm(Outstate~Private*Elite,College))
```

## Problem 9

```{r ch24problem9}
head(Auto)
apply(Auto[,1:7],2,range)
apply(Auto[,1:7],2,mean)
apply(Auto[,1:7],2,median)
apply(Auto[,1:7],2,sd)
apply(Auto[,1:7],2,mad)
apply(Auto[-(10:85),1:7],2,range)
apply(Auto[-(10:85),1:7],2,mean)
apply(Auto[-(10:85),1:7],2,sd)
pairs(Auto)
ggpairs(Auto[,-ncol(Auto)])
```
